---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# historicalborrowlong

[![check](https://github.com/wlandau/historicalborrowlong/workflows/check/badge.svg)](https://github.com/wlandau/historicalborrowlong/actions?query=workflow%3Acheck)
[![cover](https://github.com/wlandau/historicalborrowlong/workflows/cover/badge.svg)](https://github.com/wlandau/historicalborrowlong/actions?query=workflow%3Acover)
[![pkgdown](https://github.com/wlandau/historicalborrowlong/workflows/pkgdown/badge.svg)](https://github.com/wlandau/historicalborrowlong/actions?query=workflow%3Apkgdown)
[![lint](https://github.com/wlandau/historicalborrowlong/workflows/pkgdown/badge.svg)](https://github.com/wlandau/historicalborrowlong/actions?query=workflow%3Alint)

Historical borrowing in clinical trials can improve
  precision and operating characteristics. This package supports
  a longitudinal hierarchical model to borrow historical
  control data from other studies to better characterize the
  control response of the current study. It also quantifies
  the amount of borrowing through longitudinal benchmark models (independent
  and pooled). The hierarchical model approach to historical borrowing
  is discussed in
  [Viele et al. (2013)](https://onlinelibrary.wiley.com/doi/10.1002/pst.1589).

## Installation

```{r, eval = FALSE}
remotes::install_github("wlandau/historicalborrowlong")
```

## Documentation

* Functions: <https://wlandau.github.io/historicalborrowlong/reference/>
* Methods: <https://wlandau.github.io/historicalborrowlong/articles/methods.html>
* Usage: <https://wlandau.github.io/historicalborrowlong/articles/usage.html>

## Thanks

[Albert Man](https://github.com/albert-man), [Faith Bian](https://github.com/faithbian-lilly), and [Saptarshi Chatterjee](https://github.com/schatterjee-lilly) contributed to the development of the methods prior to implementation. Phebe Kemmer, Heather Zhao, and Zhangchen Zhao also provided helpful feedback on the models and their application to clinical use-cases.

## Code of Conduct

Please note that the historicalborrowlong project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
